What We Eval in the Shadows

### 2021-04-17 paper/main.tex

\subsection{Domain Specific Language.} R expressions are built and transformed
symbolically and then evalled to a concrete value. For instance, base R packages
\emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D}
and \c{deriv}. They support arithmetic operations and functions on real numbers
such as \c{sin}. In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear
Mixed Models), a \c{Dtensor} object is created, which contains expressions
derived twice (with \c{DD}) and then typically in a list or a data frame.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}

Package \c{tidyselect} is the backend of several \c{dplyr} and \c{tidyr}
functions to select columns. It offers a DSL that makes it possible to select a
set of columns, over a range of columns, a complement, an intersection, an
union, with a regular expression and so on. \eval is used to traverse selectors
stored as slots of a dataframe.

>>>>

\subsection{Domain Specific Language}

The ability to capture current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes the R toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allowing one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

Another example is the \c{tidyselect} package that defines a DSL for selecting
columns in a data frame powering number of data manipulation operation in the
\c{data.table} main competitor, the \c{dply} package. It uses eval to build an
interpreter for evaluating column selectors such as \c{select(df,
starts\_with("length") \& !(name:mass))} which returns all columns in \c{df}
that starts with the string \c{length} and is not between \c{name} and
\c{mass} columns.

For instance, base R packages \emph{stats} provides two operators for
\emph{symbolic differentiation}, \c{D} and \c{deriv}. They support arithmetic
operations and functions on real numbers such as \c{sin}. In \emph{MCMCglmm}
(Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object
is created, which contains expressions derived twice (with \c{DD}) and then
typically in a list or a data frame.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}
